How Slow do my Transient Type checks Go?

### 2019-08-17 paper.tex
Hopefully recovered from merge hell...

\begin{document}
\title{How Slow do my Transient Typechecks Go?}

@@

There is a spectrum of different approaches to gradual typing
\cite{kafka18,bensurvey18icfp}.
At one end is ``pluggable types'' (as in Strongtalk \cite{strongtalk}) or ``erasure
semantics'' (as in 
%% Two important approaches are optional\citep{}
%% and gradual typing\citep{GiladPluggable2004,Siek2006,XXXSiek2015}.
%% These are applied to dynamic languages to reap the benefits of typing, but
%% unfortunately also have limitations.
%With optional or pluggable approaches such as 
%\citeurl{TypeScript,}{TypeScript}{Microsoft}{27 June
%2018}{https://www.typescriptlang.org/}
TypeScript
\citep{typeScriptECOOP}), where
%
all types are erased before execution, limiting the benefit of
types to statically typed portions of programs, while not providing 
any for of run time checking.  In the middle there are
``transient'' or ``type-tag'' checks (as in Reticulated Python),
% and ``concrete'' checks as in Thorn
which offer first-order semantics, by checking
that an object's type constructor or supported methods match
% any explicit type annotations
any static types
that the object flows through.
%%%WRONG kjx
%, and checking only when data flows through those declarations
\cite{Siek2007,Bloom2009,concrete15,reticPython2014,Greenman2018}.
Reticulated Python also supports alternative ``monotonic'' semantics
which mutate objects by narrowing their concrete types when passed
into contexts with more specific types.
On the other end of the spectrum are behavioural
typechecks (as in Typed Racket \cite{typedScheme08,takikawa2012},
Gradualtalk \cite{gradualtalk14}),
and ``proxies'' (provided by Reticulated Python), which
support higher-order semantics, retaining
types until run 
time, and performing checks eagerly, thus detailed information
about type violations can be given as soon as possible via ``blame''
tracking \cite{blame2009,blameForAll2011}.
Finally, there is also ductile typing,
which dynamically interprets a static type system at run time
\cite{Ductile2011}.
%
%\mwh{Does a monotonic semantics fit on this spectrum somewhere?}
%\kjx{Not in the greenman/fellenstein spectrum, but I guess I'd put it
%  after concrete and before behavioural. Does it matter?}
%
% This said "any", but the ecoop paper specifically showed that 
Unfortunately, most gradual systems with run-time semantics
(as opposed to type erasure as in TypeScript) typically
imposes significant run-time performance overheads to provide these
semantics.  This has lead to a significant body of research,
on one hand developing techniques to optimise gradual typing 
\citep{Vitousek2017,Muehlboeck2017,Bauman2017,Richards2017,Greenman2018}
and on the other, developing techniques to evaluate the
resulting performance  \cite{Takikawa2016,Greenman2019jfp}.

@@

\begin{itemize}
  \item an approach to identify gradual type annotations
        that cause significant performance effects
  % SM: after discussion with Isaac, this was done
  %     by some Reticulated Python paper before
  % and Isaac's adaptation is restricted to using rounding rules
  %    (>= 0.5 is round up) that give useful results (at least 1 type check)
  %    for selecting the number of type checks
  % \item adapting the Takikawa protocol for languages and implementations
  %       where performance effects are dominated by type annotations instead
  %       of transitions between typed and untyped modules
  \item observing that performance of small benchmarks
        with 10--250 type annotations is dominated by only
        one or two type annotations
\end{itemize}

@@

To identify type annotations causing anomalies,
% To investigate Moth's support for transient type checks further, 
we adapted the Takikawa-Vitousek protocol. For each benchmark we generated 100 partially typed versions,
or fewer if the benchmark has at most 10 types.
We did an even split so that for each interval $i \ge 1$ and $i$ < $N$, we generated roughly the same number of configurations with $i$ type annotations. We used Robert Floyd's sampling algorithm \cite{Bentley:1987:PPS:30401.315746} to choose the type annotations each configuration contained randomly, and we ensured that no duplicate configurations were generated. In addition to these, we tested fully untyped and typed versions, for a total of 102 configurations per benchmark (or 97 in the case of our Storage benchmark, since it only has 10 type annotations).

@@

As an alternative, and less labour intensive approach to identifying potentially influential type annotations, for each type annotation we ran each benchmark with \textit{only} that type annotation present. We used the same computer and number of iterations as for the typing lattice experiments, albeit these were done on different days.

% We used the same computer and number of iterations as for the typing
% lattice experiments, albeit these were done on different days.

Figure \ref{f:pattern} shows the results of these experiments.
%, as well as the patterns we manually identified.
%%KJX but you haven't talked about manual patterns!

>>>>

\begin{document}
\title{How Slow do my Transient Type checks Go?}

@@

There is a spectrum of different approaches to gradual typing
\cite{kafka18,bensurvey18icfp}.
At one end is ``pluggable types'' (as in Strongtalk \cite{strongtalk}) or ``erasure
semantics'' (as in 
%% Two important approaches are optional\citep{}
%% and gradual typing\citep{GiladPluggable2004,Siek2006,XXXSiek2015}.
%% These are applied to dynamic languages to reap the benefits of typing, but
%% unfortunately also have limitations.
%With optional or pluggable approaches such as 
%\citeurl{TypeScript,}{TypeScript}{Microsoft}{27 June
%2018}{https://www.typescriptlang.org/}
TypeScript
\citep{typeScriptECOOP}), where
%
all types are erased before execution, limiting the benefit of
types to statically typed portions of programs, while not providing 
any for of run time checking.  In the middle there are
``transient'' or ``type-tag'' checks (as in Reticulated Python),
% and ``concrete'' checks as in Thorn
which offer first-order semantics, by checking
that an object's type constructor or supported methods match
% any explicit type annotations
any static types
that the object flows through.
%%%WRONG kjx
%, and checking only when data flows through those declarations
\cite{Siek2007,Bloom2009,concrete15,reticPython2014,Greenman2018}.
Reticulated Python also supports alternative ``monotonic'' semantics
which mutate objects by narrowing their concrete types when passed
into contexts with more specific types.
On the other end of the spectrum are behavioural
type checks (as in Typed Racket \cite{typedScheme08,takikawa2012},
Gradualtalk \cite{gradualtalk14}),
and ``proxies'' (provided by Reticulated Python), which
support higher-order semantics, retaining
types until run 
time, and performing checks eagerly, thus detailed information
about type violations can be given as soon as possible via ``blame''
tracking \cite{blame2009,blameForAll2011}.
Finally, there is also ductile typing,
which dynamically interprets a static type system at run time
\cite{Ductile2011}.
%
%\mwh{Does a monotonic semantics fit on this spectrum somewhere?}
%\kjx{Not in the greenman/fellenstein spectrum, but I guess I'd put it
%  after concrete and before behavioural. Does it matter?}
%
% This said "any", but the ecoop paper specifically showed that 
Unfortunately, most gradual systems with run-time semantics
(as opposed to type erasure as in TypeScript) typically
imposes significant run-time performance overheads to provide these
semantics.  This has lead to a significant body of research,
on one hand developing techniques to optimise gradual typing 
\citep{Vitousek2017,Muehlboeck2017,Bauman2017,Richards2017,Greenman2018}
and on the other, developing techniques to evaluate the
resulting performance  \cite{Takikawa2016,Greenman2019jfp}.

@@

\begin{itemize}
  \item an approach to identifying gradual type annotations
        that cause significant performance effects
  % SM: after discussion with Isaac, this was done
  %     by some Reticulated Python paper before
  % and Isaac's adaptation is restricted to using rounding rules
  %    (>= 0.5 is round up) that give useful results (at least 1 type check)
  %    for selecting the number of type checks
  % \item adapting the Takikawa protocol for languages and implementations
  %       where performance effects are dominated by type annotations instead
  %       of transitions between typed and untyped modules
  \item an observation that the overhead of Moth's transient type checking on small benchmarks (with 10--250 type annotations) is usually caused by only one or two type annotations
\end{itemize}

@@

To identify type annotations causing anomalies,
% To investigate Moth's support for transient type checks further, 
we adapted the Takikawa-Vitousek protocol. For each benchmark we generated 100 partially typed versions,
or fewer if the benchmark has less than 11 types.
We did an even split so that for each interval $i \ge 1$ and $i$ < $N$, we generated roughly the same number of configurations with $i$ type annotations. We used Robert Floyd's sampling algorithm \cite{Bentley:1987:PPS:30401.315746} to choose the type annotations each configuration contained randomly, and we ensured that no duplicate configurations were generated. In addition to these, we tested fully untyped and typed versions, for a total of 102 configurations per benchmark (or 97 in the case of our Storage benchmark, since it only has 10 type annotations).

@@

Figure \ref{f:pattern} shows the results of these experiments, together with patterns we manually identified.

The column graphs show the execution time of configurations with only a single type annotation, the x-axis indicates the index of this annotation, thus the first column represents the first annotation, and the last column represents the last one (in the order they appear in the source code). The y-axes for the column graphs are the same as the associated scatter graphs. However, though the right y-axes for the column graphs appear to be identical to the scatter plot ones, they are relative to performance results collected on different days: this was to ensure that the performance of the test computer had not changed significantly from when the typing lattice experiments were run.
